```{r setup}
library(mmsd.wq)
```

# Get Data

## get_survey_data()
The core function in mmsd.wq is **get_survey_data**.  This function
is used to query and process water quality data from the LIMS server. This is 
done by first creating ODBC connections with relevant databases 
(e.g LIMS_exchange and FreshwaterMonitoring). Once connected, **get_survey_data** 
utilizes a SQL function (fFreshwaterMonitoringOpSID) written by Kevin Moore,
along with other basic SQL queries to retrieve data. R code is used to customize
queries and to perform post-processing. 



```{r, echo = FALSE, fig.cap = "Basic Data flow of get_survey_data()"}

DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = TB]
  
  node [shape = rectangle]        
  rec1 [label = 'LIMS_Exchange']
  rec2 [label = 'fFreshwaterMonitoringOpSID']
  rec3 [label =  'get_survey_data()']
  rec4 [label = 'FreshwaterMonitoring']
  rec5 [label = 'SQL SELECT']
  rec6 [label = 'Output Data']
  rec7 [label = 'Proccess Data']
  
  # edge definitions with the node IDs
  rec1 -> rec2 -> rec3
  rec4 -> rec5 -> rec3
  rec3 -> rec7
  rec7 -> rec6
  }",
  height = 500)


```


## ODBC Connections

The *get_survey_data* functions relies on ODBC connections to retrieve
data from databases. A connection can be made using the *make_dBConnection()*
function. 

```{r, eval = FALSE}
#make a connection to the "LIMS_Exchange" database
LIMS <- make_dBConnection(database = "LIMS")

#make a connection to the "FreshwaterMontoring" database
freshwater <- make_dBConnection(database = "freshwater")

```

Once connections are made, they can be passed to the *get_survey_data* arguments
to retrieve data. Both a connection to the "LIMS_Exchange" and 
"FreshwaterMonitoring" databases are required. 

```{r, eval = FALSE}

wq_data <- get_survey_data(LIMS_con = "LIMS",fw_con = "freshwater",
                           survey_num = "SS 400")

```

It is best practice to close ODBC connections at the end of every session. This
can be done using the *odbc* package.

```{r, eval = FALSE}

odbc::dbDisconnect(freshwater)
odbc::dbDisconnect(LIMS)

```

If database connections are not provided to the *get_survey_function* these
will be made internally within the function. These connections will be
closed after making all necessary queries.  As a check, messages will be printed
indicating that these connections were made then closed. 

```{r}

wq_data <- get_survey_data(survey_num = "SS 400")


```

## Output Data

The output of the get_survey_data is a dataframe containing water
quality data and it's associated meta data. The function works by joining 
the raw query output returned from the "fFreshwaterMonitoringOpSID" SQL 
function to internal package lookup tables and tables from SQL databases.

```{r, echo = FALSE}

DT::datatable(wq_data, options = list(
  pageLength=5, scrollX='400px'))
```

## Output Data Definitions 

```{r, echo = FALSE}

tibble::tribble(
  ~field,~description,
  "sample_sid", "The unique sample id associated with the result",
  "collection_date","The date the sample was collected",
  "collection_time","The time the sample was collected",
  "collection_datetime","The data and time the sample was collected",
  "year","The year the sample was collected",
  "month","The month the sample was collected",
  "site_code","The site code and depth formatted like 'OH-01S'",
  "sitecode","The site code without depth formatted like 'OH-01'",
  "depth", "The depth at which a sample was collected ('S','M','B')",
  "latitude", "The latitude associated with the site",
  "longitude", "The logitude assocaited with the site",
  "site_no","A unique site identifer",
  "site_type", "The site category ('River','Lake','Outfall','Harbor')",
  "survey_num", "The unique ID for the given survey",
  "source_code_sid", "The source code identifying the survey type",
  "survey_type","Name associated with source code",
  "survey_group","A name that groups survey types together by when they are completed (e.g. Quad Creek)",
  "water_body","The water body that the site belongs to (e.g. Kinnickinnic River)",
  "op_sid", "The unique identifier that designates a water quality parameter. Short for Operations ID",
  "label_name","A short, clean name for a water quality parameter",
  "parameter","The full name for a water quality parameter",
  "reading_val","The result for the sample formatted as a character",
  "reading_num","The result for the sample formatted as a number.  This is typically the result used by FRM staff",
  "reading_raw","The result for the sample reported by the lab equipment",
  "hours_dry","The number of hours since the sample site recieved precipiation. Calculated from HACH WIMS",
  "precip72hr","The inches of precipitation the site has recieved in the last 72 hours. Calculated from HACH WIMS",
  "high_limit","The upper limit for the site-parameter combo calculated within sample review",
  "low_limit","The low limit for the site-parameter combo calculated within sample review",
  "loq", "The Limit of Quantitation" ,
  "mdl", "The minimum quantification limit",
  "reading_flag","A binary variable",
  "result_flag", "Lab flags associated with the sample ",
  "samplte_type","Orignal, field blank, or field replicate sample",
  "qa_code_sid", "A binary variable identifying if a sample is Q flagged",
  "q_flag_description","Any Q flags associated with the sample",
  "result_entered_datetime", "The date and time the sample was entered into LIMS",
  "result_comments","Any general comments associated with the sample"
  ) %>%
  DT::datatable(options = list(
  pageLength=5, scrollX='400px'))


```


### Wide Data

By default, data are returned in long format.  If preferred, data can be returned
in wide format by specifying **wide = TRUE**.  By the nature of the data,
some columns will be dropped from the output if the data are formatted wide.

```{r}

wq_data_wide <- get_survey_data(survey_num = "SS 400", wide = TRUE)

```


```{r, echo = FALSE}
DT::datatable(head(wq_data_wide),options = list(
  pageLength=5, scrollX='400px'))
```


### Raw Data

By setting "raw = TRUE" a user can bypass any post-proccessing and output the
raw results returned from the "fFreshwaterMonitoring" SQL function. 


```{r}

wq_data_raw <- get_survey_data(survey_num = "SS 400", raw = TRUE)

```


```{r, echo = FALSE}
DT::datatable(head(wq_data_raw),options = list(
  pageLength=5, scrollX='400px'))
```

## Query Parameters

*get_survey_data* has several arguments that can be used to refine a 
query. A "start_date" and "end_date", "survey_num", or sample_sid  
are required.  All other arguments are optional.  Multiple arguments can be 
used to create a specific query. Built in error checking code limits the user 
from creating nonsensical queries. The following sections show how to use the 
different arguments to refine queries. 

### Date Range

The "start_date" and "end_date" arguments can be used to retrieve data
within a certain date range.  Dates need to be passed as a character object
and need to be formatted as "YYYY-MM-DD" (e.g. "2020-10-31").  


```{r, eval = FALSE}
#get all data collected in  2020
wq_data <- get_survey_data(start_date = "2020-01-01",
                           end_date = "2020-12-31")

 
```

### Survey Number

Data for an individual survey can be retrieved by using the *survey_num* 
argument. The *survey_num*  must be a character string with the 
survey type (e.g. "OH", "SS", "RR") and it's unique survey number separated 
by a space. Multiple surveys can be retrieved by concetanting their unique
survey identifiers. 

```{r, eval = FALSE}

#get all data for SS 400
wq_data <- get_survey_data(survey_num = "SS 400")

#get data from multiple surveys
wq_data <- get_survey_data(survey_num = c("SS 400","SS 401", "SS 402"))

```

### Sample SID
The sample_sid argument can be used to get data for a specific Sample SID(s).
Sample SID's need to be passed to the arguement as a character object 
(e.g. "20003678" ).  Multiple Sample SID's can be passed during a single
query.

```{r, eval = FALSE}

sample <- get_survey_data(sample_sid = "20003678")
samples <- get_survey_data(sample_sid = c("20003678","20003679"))

```

### Sites

Data can be retrieved for specific sites using the *site_code* or *sitecode*
arguments.  The full site identifier, including depth, can be passed to the
*site_code* argument (e.g. "SS-01S","OH-14B").  Conversely, the site identifier 
without depth can be passed to the sitecode argument to retrieve all depths
associated with that site.  For example, passing "OH-01" will retrieve data
for "OH-01S", "OH-01M", and "OH-01B".   Multiple site identifiers can be 
passed to either argument to retrieve data for more than one site. 

```{r, eval =  FALSE}
#get data for site_code "SS-01S" for a given date range
SS_01S <- get_survey_data(start_date = "2020-01-01", end_date = "2020-12-31",
                          site_code = "SS-01S")

#get data for sitecode "SS-01".  Data for all three site depths will be
#returned. 
SS_01 <- get_survey_data(start_date = "2020-01-01", end_date = "2020-12-31",
                          sitecode = "SS-01")


#get data for SS and OH outfalls. 
outfall_data <- get_survey_data(
  start_date = "2020-01-01",
  end_date = "2020-12-31",
  sitecode = c("SS-01", "OH-02")
)

```

### Site Category

Sites are placed into four different categories by FRM staff.  These 
categories include "River", "Outfall", "Lake", and "Harbor".  Passing
any of these values to the "site_category" argument will retrieve sites
with the relevant attribute. 

```{r, eval = FALSE}

#get 2020 outfall data
outfall2020 <- get_survey_data(start_date = "2020-01-01",
                             end_date = "2020-12-31",
                             site_category = "Outfall")


#get 2020 lake and harbor data
lake_michigan_2020 <- get_survey_data(start_date = "2020-01-01",
                             end_date = "2020-12-31",
                             site_category = c("Lake","River"))

```


### Source Code

Data for given source codes can be retrieved using the *source_code* argument.
Available source codes with their site codes can be found using the internal 
lookup table site_code_lookup. Multiple values can be 
passed to retrieve data for more than one source_code.  

```{r, eval = FALSE}

#get 2020 outer harbor data
oh2020 <- get_survey_data(start_date = "2020-01-01", end_date = "2020-12-31",
                          source_code = 900)

#get 2020 Lake Michigan Data (OH,SS,NS)
oh2020 <- get_survey_data(start_date = "2020-01-01", end_date = "2020-12-31",
                          source_code = c(900,902,903))


```

###  WQ Parameters

Data can be retrieved for specific parameters using the *op_sid* or 
*param_name* arguments. An operation ID(s) can be passed to the *op_sid* 
argument.  Available operation ID's can be found in the internal lookup
table "param_lookup_table".  Parameter names can be passed to the *param_name*
argument.  The internal vector "parameter_names" can be used to see the 
available arguments for "param_name". Multiple values can be 
passed to either argument to retrieve data for more than one parameter. 

```{r, eval = FALSE}
#get DO data by operations ID for survey SS 400
DO_data <- get_survey_data(survey_num = "SS 400", op_sid = 182)

#get 2020 turbidity data
turb2020 <- get_survey_data(start_date = "2020-01-01", end_date = "2020-12-31",
                           param_name = "Turbid")

#get 2020 bacteria data 
bac2020 <- get_survey_data(start_date = "2020-01-01", end_date = "2020-12-31",
                           param_name = c("E_coli","FCMF"))

```


### Sample Depth

The *collection_depth* argument can be used to retrieve data from a specified 
depth only.  Options include "S","M","B".  Multiple depths can be specified.

```{r, eval = FALSE}

#Collect all mid depth data for survey SS 400
surface_data <- get_survey_data(survey_num = "SS 400", collection_depth = "M")


```


### Blanks and Replicates

By default blanks and replicates are not included in output data.  
The arguments *blanks* and/or *reps* can be set to TRUE include either. Likewise,
the argument orig_samples can be set to FALSE to exclude original samples
(non QA/QC). 

```{r, eval = FALSE}

#get the blanks and replicates for survey SS 400
blanks_reps <- get_survey_data(survey_num = "SS 400", blanks = TRUE,
                               reps = TRUE, orig_samples = FALSE)


```

### Flags

Q and M flags are not returned by default.  To return these results,
m_flag and/or q_flag can be set to TRUE.

```{r, eval = FALSE}

#get all 2020 water quality data - M and Q flags included, 
flags_included <- get_survey_data(start_date = "2020-01-01", 
                                  end_date = "2020-12-31",
                                  q_flag = TRUE,
                                  m_flag = TRUE)

```

### LOQ

The above_LOQ argument can be set to TRUE to only return data that is above the
LOQ. This argument is set to FALSE by default. 

```{r, eval = TRUE}

above_LOQ <- get_survey_data(start_date = "2020-01-01", 
                             end_date = "2020-12-31",
                             above_LOQ = TRUE)


```


## Transformations and Calculations

### Data Transformations

The *transformations* argument can be used to apply a mathematical 
transformation to the reading_num column.  Currently, "log10" and "log"
transformations are available. 

```{r,eval = FALSE}

wq_data_log10 <- get_survey_data(survey_num = "SS 400", 
                                 transformation = "log10")

wq_data_log <- get_survey_data(survey_num = "SS 400", 
                                 transformation = "log")
                               
```

### Outliers

Outliers in the output data can be identified by using the *calculate_outliers*
argument. Methods include *quantile* and *sd*.  The *quantile* method identifies
mild and extreme outliers using methods outline in SOP 035.  The *sd* method
if the result is 2 times the standard deviation.  If this argument is used
a column will be added to the output data identifying which results are 
outliers. 

```{r, eval = FALSE}

wq_data_quant <- get_survey_data(survey_num = "SS 400", param_name = "DO",
                                 calculate_outliers = "quantile")

wq_data_quant <- get_survey_data(survey_num = "SS 400", param_name = "DO",
                                 calculate_outliers = "sd")


```

### Handling Non-Detects
When non-detects (ND's) are reported the value in the *reading_num* field can be
returned in four different ways.  
* "zero" - ND's are returned as zeros
* "half" - ND's are returned as half of their MDL.
* "MDL" - ND's are returned as the MDL.
* "remove" - ND's are no returned by the function. 
Any of these options can be passed to the *MDL_option* argument.  The default
value for this argument is "half".

```{r, eval = FALSE}

wq_data <- get_survey_data(start_date = "2020-01-01", end_date = "2020-12-31",
                           MDL_option = "zero")

```


